---
title: "hw1"
author: "James Dalgleish"
date: "September 13, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Here, we're creating a tibble, essentially a data.frame, with multiple data types, numeric, logical, character, and factor making up four distinct columns.
what didn't work:
letters(1:5)
"letters" is a built in vector rather than a function.
```{r create_tibble}
rand10_greaterthan2_letters_factor_tibl <- tibble::tibble( 
  ten_random_0_to_5 = runif(n = 10, min = 0, max=5),
  logical_greater_than_two = ten_random_0_to_5 > 2,
  first_ten_letters = letters[ 1:10 ],
greater_than_two_factor= as.factor(logical_greater_than_two)

)
head(rand10_greaterthan2_letters_factor_tibl)
```
Now, we'll attempt to take the mean of all the columns... we'll note that some
of these columns are
not numeric and therefore a mean would be difficult to calculate.
```{r means}
#taking the mean of every column won't work because 
#you can't take the mean of a nonnumeric column.
 tibl_means <- sapply( rand10_greaterthan2_letters_factor_tibl, mean)
mean( rand10_greaterthan2_letters_factor_tibl$ten_random_0_to_5 )
mean( rand10_greaterthan2_letters_factor_tibl$logical_greater_than_two )
mean( rand10_greaterthan2_letters_factor_tibl$first_ten_letters )
mean( rand10_greaterthan2_letters_factor_tibl$greater_than_two_factor )
#print(tibl_means)
```
You can actually take the mean of a factor, but it's something to avoid. It is
possible to have a factor that looks like a list of numbers that is not
a numeric variable, Using some if logic, we find we can take just the means
of the columns that are amenable to means.

```{r means_with_check}
 #putting in a check to determine if it's numeric.
tibl_means <- sapply( rand10_greaterthan2_letters_factor_tibl,
                     function(x)
                     if ( is.numeric( x ) ) {
                     return( 
                       mean( x )
                       )
                     } else {
                     return( "Incompatible type for mean" )
                     })
 print(tibl_means)
```
Converting the logical, character, and factor columns to numeric:
```{r,eval=F}
tibl_as_num_log_char_num <-
  sapply( rand10_greaterthan2_letters_factor_tibl,
  function(x)
  if (is.logical(x) |
  is.character(x) |
  is.factor(x)) {
  return(as.numeric(x))
  } else {
  return(NULL)
  })
  print( tibl_as_num_log_char_num )

```
The logical TRUEs have been converted to 1s and the logical 0s have been
converted to 
