---
title: "hw1"
author: "James Dalgleish"
date: "September 13, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```
#Problem 1
Here, we're creating a tibble (essentially a data.frame) with multiple data
types (numeric, logical, character, and factor) making up four distinct columns.

```{r create_tibble}
rand10_greaterthan2_letters_factor_tibl <- tibble::tibble( 
  ten_random_0_to_5 = runif(n = 10, min = 0, max=5),
  logical_greater_than_two = ten_random_0_to_5 > 2,
  first_ten_letters = letters[ 1:10 ],
greater_than_two_factor = as.factor( logical_greater_than_two )

)
head(rand10_greaterthan2_letters_factor_tibl)
```
Now, we'll attempt to take the mean of all the columns one by one...
we'll note that some of these columns are not numeric and therefore a mean 
would be difficult to calculate. Taking the mean of every column
won't work because means are only meaingful on numeric 
and integer vectors.
Please note: sapply() will apply the mean() function to each column in the
tibble, from the first column to the last. In this case, two of these attempts
work while the last two fail as the type is inappropriate for taking a mean.

```{r means}
 tibl_means <- sapply( rand10_greaterthan2_letters_factor_tibl, mean)
print( tibl_means )
mean( rand10_greaterthan2_letters_factor_tibl$ten_random_0_to_5 )

```
*What works?* Logical vectors.
You can also take the mean of a logical TRUE/FALSE, with TRUE being
automatically converted to 1 and FALSE to zero. This result may not have a 
straightforward interpretation, but it is possible.

```{r mean_logical}
mean( rand10_greaterthan2_letters_factor_tibl$logical_greater_than_two )

```
*What doesn't?* Character vectors.
*Why?* Taking the mean of a character vector doesn't make sense
and doesn't work. A mean is a mathematical operation that works only on numbers
or something that is automatically coerced to numbers.
See below.
```{r mean_letters}
mean( rand10_greaterthan2_letters_factor_tibl$first_ten_letters )

```
*What doesn't?* Factors, in this case.
*Why?* You cannot take the mean of a factor. The error message says it all...
it's best if you have a numeric or logical vector here. If you meant to take a 
mean by coercion from character to logical to numeric, you could using the below
code. Interpretation is not obvious, but it signifies the proportion of TRUE
values in this case (6 out of 10 are TRUE). 

```{r mean_factor}
mean( rand10_greaterthan2_letters_factor_tibl$greater_than_two_factor )
```
```{r mean_factor_coercion}
rand10_greaterthan2_letters_factor_tibl$greater_than_two_factor %>%
  as.character() %>% 
  as.logical() %>% 
  as.numeric() %>% 
  mean()
```

A practical application of noticing that some types are appropriate for means:
Using some conditional logic, we find we can take just the means
of the columns that are amenable to means.

```{r means_with_check}
 #putting in a check to determine if it's numeric.
tibl_means <- sapply( rand10_greaterthan2_letters_factor_tibl,
                     function( x )
                     if ( 
                       is.numeric( x )
                       ) {
                     return( 
                       mean( x )
                       )
                     } else {
                     return( "Incompatible type for mean" )
                     })
 print(tibl_means)
```
We now convert the logical, character, and factor columns to numeric.
The results are hidden, as per instructions. If you wish to view the results to
verify the function works, one can choose to print the results with the provided
print function call.
```{r converting_to_numeric,results = "hide"}
tibl_as_num_log_char_num <-
  sapply( rand10_greaterthan2_letters_factor_tibl,
  function( x )
  if (is.logical( x ) |
  is.character( x ) |
  is.factor( x ) ) {
  return( as.numeric( x ) )
  } else {
  return(NULL)
  })
  print( tibl_as_num_log_char_num )

```
*What works?*
If you were to print the results,the logical TRUEs have been converted to 1s
and the logical FALSEs have been converted to 0s.
*What doesn't?*
The letters are completely incompatible and result with 
empty values (NAs).
*Why?* The logical vector can be converted to numeric,
so this is something to watch out for and avoid if not intended.

##Converting character variable from character to factor to numeric:
```{r char_to_fact_to_num}
print( "vector class:" )
class( rand10_greaterthan2_letters_factor_tibl$first_ten_letters )
print( "character to vector:" )
as.factor( rand10_greaterthan2_letters_factor_tibl$first_ten_letters ) 
print( "character to vector to numeric:" )
as.factor( rand10_greaterthan2_letters_factor_tibl$first_ten_letters ) %>% 
  as.numeric()
```
*What happens?*
Converting a 10 length character vector to factor generates a factor with
10 levels of length 10. Converting this factor to numeric gives the number 
(index) corresponding with the factor level at each position.

-----
##Converting a factor to character to numeric.
```{r fact_to_char_to_num}
rand10_greaterthan2_letters_factor_tibl$greater_than_two_factor
as.character( rand10_greaterthan2_letters_factor_tibl$greater_than_two_factor )
as.numeric( 
  as.character( rand10_greaterthan2_letters_factor_tibl$greater_than_two_factor )
)
```
The factor levels are appropriately converted from TRUE to "TRUE", but R doesn't
quite know how to convert "TRUE" into a number. Each of the attempted
conversions turn into missing values (NAs).

#Problem 2
Following directions from the assignment, I will create a data frame that has
the following vectors as columns:
*two numeric variables, x and y,
**both taken from a standard normal distribution
**(standard signifies mean 0, sd 1),
*logical vector denoting x + y > 0
*the logical vector coerced into a numeric vector
*the logical vector coerced into a factor vector


```{r create_second_tibble}
tibl_rnorm_1k_x_y_gt_0_num_fac <- tibble::tibble(
  x = rnorm( n = 1000, mean = 0, sd = 1 ),
  y = rnorm( n = 1000, mean = 0, sd = 1 ),
  gt_0 = x + y > 0,
  gt_0_num = as.numeric( gt_0 ),
  gt_0_fac = as.factor( gt_0 )
)
head( tibl_rnorm_1k_x_y_gt_0_num_fac )
```

To illustrate the size of the dataset, the following code will show that there
are `r nrow( tibl_rnorm_1k_x_y_gt_0_num_fac )` rows and
`r ncol( tibl_rnorm_1k_x_y_gt_0_num_fac )` columns.
```{r second_tibble_size}
print("number of rows in tibble:" )
nrow( tibl_rnorm_1k_x_y_gt_0_num_fac )
print( "number of columns in tibble:" )
ncol( tibl_rnorm_1k_x_y_gt_0_num_fac )
```
Showing the mean and median of the variable x is also simple using the functions
bearing those precise names:
```{r x_mean}
mean( tibl_rnorm_1k_x_y_gt_0_num_fac$x )
```

```{r x_median}
median( tibl_rnorm_1k_x_y_gt_0_num_fac$x )

```

To find the proportion of TRUE cases in the logical vector, we might do the 
following using the table and print commands, with some basic math:
```{r true_prop_cases_logical}
n_true <- table( tibl_rnorm_1k_x_y_gt_0_num_fac$gt_0 ) [ "TRUE" ]
n_false <- table( tibl_rnorm_1k_x_y_gt_0_num_fac$gt_0 ) [ "FALSE" ]
prop_true <- n_true / ( n_true + n_false )
print( "number true:" )
print( n_true )
print( "number false:" )
print( n_false )
print( "proportion true:" )
print( prop_true )
```
Here's a scatterplot of Y vs X, with the colors set by the logical vector in
the data frame (or tibble, however you'd like to call it).
FYI: I do in fact know how to use library function, but I prefer to use "::"
notation to be in habits avoid package namespace conflicts.
I've opted for an interactive plot here, using the plotly conversion function
instead of the print() function. Exporting the ggplot object using ggsave()
also happens below.
The variable is used so that "tibl_rnorm_1k_x_y_gt_0_num_fac$gt_0_fac"
doesn't end up in labeling.
```{r first_scatterplot_and_save}
greater_than_zero <- tibl_rnorm_1k_x_y_gt_0_num_fac$gt_0_fac
y_vs_x_colored_by_logical <- ggplot2::ggplot( data = 
                tibl_rnorm_1k_x_y_gt_0_num_fac,
                ggplot2::aes(x = x,
                y = y,
                color = greater_than_zero)) +
                ggplot2::geom_point() +
                ggplot2::labs(color = "x + y > 0")

  plotly::ggplotly( y_vs_x_colored_by_logical )
 ggplot2::ggsave( "y_vs_x_colored_by_logical.pdf", 
          plot = y_vs_x_colored_by_logical,
          device = "pdf")
```
Here is another plot with the plot colored with a numeric variable.
Notice the continous color scale denoted by the gradient bar.
There really are still just two colors for 0 and 1, however.
```{r ggplot_with_num_vars}
greater_than_zero <- tibl_rnorm_1k_x_y_gt_0_num_fac$gt_0_num
y_vs_x_colored_by_num <- ggplot2::ggplot(data = 
                tibl_rnorm_1k_x_y_gt_0_num_fac,
                ggplot2::aes(x = x,
                y = y,
                color = greater_than_zero)) +
                ggplot2::geom_point() +
                ggplot2::labs(color = "x + y > 0")
  plotly::ggplotly( y_vs_x_colored_by_num )
```
Here is another plot with the plot colored with a factor variable. Notice the discrete color scaling, just like the logical plot (the first one).
```{r ggplot_with_factor_vars}
greater_than_zero <- tibl_rnorm_1k_x_y_gt_0_num_fac$gt_0_fac
y_vs_x_colored_by_factor <- ggplot2::ggplot( data = 
                tibl_rnorm_1k_x_y_gt_0_num_fac,
                ggplot2::aes( x = x,
                y = y,
                color = greater_than_zero)) +
                ggplot2::geom_point() +
                ggplot2::labs(color = "x + y > 0" )
  plotly::ggplotly( y_vs_x_colored_by_factor )
```